\documentclass[11pt,a4paper]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{hyperref}
\usepackage[margin=1in]{geometry}
\usepackage{natbib}
\usepackage{xcolor}
\usepackage{enumitem}
\usepackage{caption}
\usepackage{array}
\usepackage{tabularx}
\usepackage{multirow}
\usepackage{tikz}
\usetikzlibrary{arrows.meta, positioning, calc, fit, backgrounds,
                shapes.geometric, decorations.pathreplacing, decorations.pathmorphing}
% All figures in this draft use manual TikZ positioning, so we stay on pdfLaTeX.
% To switch to auto-laid-out figures (Fig. 2, Fig. 4), uncomment the two lines
% below AND compile with LuaLaTeX (lualatex main.tex). See figure-source comments.
% \usetikzlibrary{graphs,graphdrawing}
% \usegdlibrary{layered,force}
\usepackage{pgfplots}
\pgfplotsset{compat=1.18}

\input{notation}

% Give figures more breathing room
\setlength{\textfloatsep}{20pt plus 4pt minus 4pt}
\setlength{\intextsep}{18pt plus 4pt minus 4pt}
\setlength{\floatsep}{18pt plus 4pt minus 4pt}
\setlength{\abovecaptionskip}{10pt}
\setlength{\belowcaptionskip}{6pt}

\newtheorem{definition}{Definition}[section]
\newtheorem{proposition}[definition]{Proposition}

\title{\textbf{Norms as Shared Precision Priors: \\[0.3em]
Values, Salience, and Cultural Transmission \\[0.3em]
in Active Inference}}
\author{Jonas Hallgren\textsuperscript{1} \\
\textsuperscript{1}Equilibria Network \\
\texttt{contact@eq-network.org}}
\date{\today \\ \textit{Working Draft --- v0.1}}

\begin{document}

\maketitle

\begin{abstract}
Norms --- shared regularities of behaviour that enable coordination among
agents with heterogeneous preferences --- are studied separately across game
theory, cultural evolution, cognitive sociology, and active inference, with
limited integration. The substrate problem is how prior preferences come to be
shared across a community in the first place: how values are installed in
individual agents through community-level dynamics. We address this problem
within the active-inference framework. Values are identified with prior
preferences ($\C$) in an agent's generative model; prestige cues and
credibility-enhancing displays are formalised as precision modulators on the
channel through which $\C$ is updated from observed behaviour; and norms are
characterised as stabilised regions of high-precision shared $\C$ across a
community. The resulting fixed-point dynamics --- community priors shape
salience, salience directs attention and copying, copying installs priors in
next-generation agents --- admit three regimes (convergence on a shared norm,
bifurcation into multiple sub-community attractors, and dissolution into
incoherence) determined by the precision gain of the social channel. The
shape of $\C$ itself further governs the exploration--exploitation tradeoff
in installed values. The account is complementary to existing treatments of
social influence as policy-precision modulation, operating one layer up on
the content of prior preferences. We propose a multi-agent simulation that
operationalises the regime predictions and discuss implications for tracking
how mediated information environments may restructure community-level value
distributions.
\end{abstract}

%% ============================================================
\section{Introduction}
\label{sec:intro}
%% ============================================================

The cognitive and cultural substrates of norms --- shared regularities of
behaviour through which agents with heterogeneous preferences coordinate
---  have been the subject of sustained inquiry across multiple disciplines,
with limited integration. Game theory and the study of conventions, descended
from Lewis, frame norms as equilibria of repeated interaction
\citep{koster2022SillyRules,oldenburgZhiXuan2024BayesianRule,
vinitsky2023sanctions}. Cultural evolution
\citep{henrichGilWhite2001Prestige,heyes2018CognitiveGadgets} documents the
population-level mechanisms by which behaviours and values are transmitted
across generations, with particular attention to prestige-biased copying and
credibility-enhancing displays \citep{henrich2009CREDs,chudek2012BystanderGaze}.
Cognitive sociology, anchored in Bourdieu's account of habitus and developed
empirically by \citet{vaisey2009DualProcess}, locates norms in durable
dispositions acquired through participation in social life. Active inference
\citep{friston2015ActiveInferenceEpistemic,parrFriston2017Attention} provides
a computational vocabulary in which values are prior preferences in an
agent's generative model and attention is precision-weighting on prediction
error; multi-agent extensions of this framework
\citep{albarracin2022EpistemicCommunities,heins2024CollectiveSurprise} have
begun to apply these tools to collective phenomena.

What has been missing is a formal account of the substrate that connects
these traditions: the cognitive mechanism by which the prior preferences
that make a norm binding for an individual agent come to be shared across a
community. Equilibrium framings yield stable behaviour without specifying a
cognitive substrate; cultural evolution describes the empirical transmission
process in informal vocabulary; cognitive sociology characterises the
phenomenology without committing to a computational mechanism; active
inference has the formal apparatus but has not been systematically applied
at the junction where individual generative models are coupled by social
observation. The substrate problem is the gap into which this paper places
its contribution.

Working in the active-inference framework, we identify values with prior
preferences ($\C$) in an agent's generative model. We propose that prestige
cues and credibility-enhancing displays modulate the precision of the
channel through which $\C$ is updated from observed behaviour, and that
norms are best characterised as stabilised regions of high-precision shared
$\C$ across a community. The resulting fixed-point loop --- community priors
shape salience, salience directs attention and copying, copying installs
priors in next-generation agents --- admits three regimes (convergence on a
shared norm, bifurcation into echo chambers, dissolution into incoherence)
determined by the precision gain of the social channel. The shape of $\C$
itself --- sharp, broad, or what \citet{torresan2025ShapeC} call ``shaped''
--- governs the exploration--exploitation tradeoff and the robustness of
installed values.

The contribution is threefold. First, we provide a formal definition of
norms as high-precision shared regions of $\C$, with explicit stability
conditions for the loop dynamics. Second, we differentiate the account from
\citet{albarracin2025SocialPower}'s treatment of social power as
policy-precision modulation: we operate one layer up, on the content of
prior preferences. Third, we propose a minimum viable multi-agent simulation
that operationalises the regime predictions and provides a target for
empirical work.

The paper proceeds as follows. Section~\ref{sec:cooperative-ai} situates the
substrate problem within contemporary norm research in multi-agent systems
and cognitive science. Section~\ref{sec:cultural-evolution} surveys the
cultural-evolution backdrop, paying particular attention to the
domain-generality of prestige cues, which behave as precision parameters
rather than content priors. Section~\ref{sec:formal} develops the formal
account, including the definition of norms and the bifurcation conditions
of the loop. Section~\ref{sec:simulation} proposes a multi-agent simulation
that operationalises the framework's regime predictions.
Section~\ref{sec:discussion} closes with limitations and remarks on the
implications for environments in which artificial agents mediate the
substrate of value transmission.

\input{section_1}

\input{section_2}

\input{section_3}

\input{section_4}

%% ============================================================
\section{Discussion and Open Work}
\label{sec:discussion}
%% ============================================================

We have argued that cooperative AI's working concept of norm is best refined
as a stabilised region of high-precision shared prior preference, installed
in agents by a precision-weighted attentional channel that prestige cues and
credibility-enhancing displays modulate. The fixed-point loop we formalise
--- community priors shape salience, salience directs copying, copying
installs priors --- is the same loop that the cultural-evolution tradition
documents informally and that active inference has the apparatus to render
formally; the contribution is to close the loop explicitly and to locate it
on the prior-preference content layer ($\C$), one level up from the
policy-precision account of social power developed by
\citet{albarracin2025SocialPower}.

Three threads warrant brief comment.

\paragraph{Gradual disempowerment as channel capture.}
The substrate of normative life, on this account, is the precision channel
through which $\C$ is installed; whoever mediates that channel shapes which
priors get installed in the next generation. The
\citet{kulveit2025GradualDisempowerment} thesis acquires a clean mechanistic
reading in this picture: the worry is not that AI systems will explicitly
override human preferences but that they already mediate the channel through
which preferences are formed. Recommendation systems, conversational
interfaces, and AI-shaped attention markets are precision-allocators, not
just content-providers. Whether their integration constitutes capture,
augmentation, or genuine partnership turns on whether the
$\C$-installation channel remains a function of the community's own values
or becomes a function of the system's optimisation objective. We flag this
as a first-order question for cooperative AI without developing it here.

\paragraph{Limitations and what is not in this paper.}
Several formal moves require empirical anchoring we do not provide. The
CRED-as-precision-modulator hypothesis is novel and predicts specific
dose-response relationships between display costliness and posterior
precision on transmitted values; the simulation in
Section~\ref{sec:simulation} tests one consequence (regime transitions in
prestige-precision gain) but not this empirical specification directly. A
follow-up empirical programme would profitably target precision-shift
markers under controlled CRED exposure. Separately, this paper does not
develop the structural / spectral reading hinted at in
Section~\ref{sec:simulation}'s in-community measurements: the broader
project of using spectral-graph quantities as cross-mechanism signatures
of coordination structure is left for a separate paper. Finally, the
political-theory bridge (Bourdieu's habitus, Lukes' third face of power as
preference-shaping) is invoked as motivation but not formalised here beyond
the explicit differentiation from \citet{albarracin2025SocialPower}.

\paragraph{What we hope this draft does.}
We have presented this as a working draft for active-inference colleagues.
We welcome feedback, especially from those embedded in multi-agent
active-inference, cultural-evolution, or cooperative-AI norm research,
where the framework most directly intersects existing programmes and where
the formalism is most likely to over- or under-reach.

\bibliographystyle{plainnat}
\bibliography{values-salience}

\end{document}
