% =====================================================================
\section{Introduction}
\label{sec:intro}
% =====================================================================

Scientific communities do not revise their theoretical commitments the
moment the evidence turns. Anomalies accumulate; the prevailing
framework persists; the shift, when it comes, is abrupt and
path-dependent \citep{kuhn1962Structure}. The persistence of Newtonian
mechanics well into the era of measurable relativistic corrections is
the canonical example: most experiments simply did not probe the
regime in which the theories disagree, and even once they did, the
community moved slowly and unevenly. The interesting question is not
why belief lags --- a Bayesian agent with a sharp prior is the obvious
proximate cause --- but \emph{what generates the gradient against
which belief is moving}. Our reading, following the sociology of
science \citep{bourdieu1975ScientificField} and the empirical
science-of-science literature \citep{wang2021ScienceOfScience,
azoulay2019StarsShadow}, is that the gradient is material: scientific
communities are \emph{coupled belief--resource systems}, and beliefs
are sustained not only by their predictive power but by the resource
topology that rewards some belief moves and punishes others.

This paper formalises that reading as a multi-agent active-inference
model with resource flow at the substrate level. Each agent
\citep{friston2015ActiveInferenceEpistemic} maintains a Gaussian
posterior over a parameterised theory and chooses experiments to
maximise expected resource gain; agents are coupled on a structured
social graph \citep{wattsStrogatz1998SmallWorld,
albarracin2022EpistemicCommunities}, with a learned channel precision
--- the \emph{trust matrix} $\trustmat$, updated by Gamma-conjugate
Bayesian precision learning on cross-surprisals --- as the only
relational learned object in the model. The substrate adds a scalar
resource state $\resource{i}$ per agent that is a \emph{deterministic
readout of $\trustmat$}: the patronage flow $\flowmat$ is the
row-normalised trust matrix, and the exogenous inflow is allocated by
aggregate incoming trust. Information gain enters the policy
\emph{instrumentally}, not as a separate intrinsic value: discriminating-%
regime experiments are the only ones whose cross-surprisals
differentiate agents and reshuffle trust, which is the only route
through which an agent earns resources. A per-agent cognitive trait
$\stub$ scales a socially-induced affective utility in the sense of
\citet{hylandAlbarracin2025MindChange} that pulls the posterior
toward the beliefs of resource-rich neighbours. The result is two
closed feedback loops --- one for trust, one for resources --- sharing
one world--agent--policy--action spine, with paradigm persistence as
the substrate-level dynamics of a population of agents who are
simultaneously investigating the world and surviving in the field.

The interesting dynamics of paradigm change --- contested periods,
factions that convert at different rates, a mainstream held back
while a vanguard moves --- are properties of population
\emph{structure} and of the \emph{distribution} of resources across
agents, not of their averages. A homogeneous population on a symmetric
graph is a degenerate case: trust-weighted pooling collapses to a
uniform neighbour average, the population moves as a single pooled
trajectory, and there is nothing for the social or resource structure
to do. We take a structured, heterogeneous population as the canonical
object and report per-community and per-agent quantities throughout
--- supplemented by \emph{resource-weighted} population summaries that
capture which agents' beliefs are materially consequential --- never a
single unweighted mean.

The contribution is a model and an experimental programme.
Section~\ref{sec:related} situates the work against its immediate
neighbours in active inference, the sociology of science, and progress
studies. Section~\ref{sec:formal} develops the formal background: the
parameterised reality, per-agent conjugate inference, trust as a
learned channel precision (the central Bayesian learned object), the
resource topology read off the trust matrix, and the cost of
mind-change as a per-agent cognitive trait.
Section~\ref{sec:model} specifies the coupled multi-agent dynamics and
states the trust-driven path-dependence claim the programme targets.
Section~\ref{sec:experiments} sets out a three-axis experimental
programme --- initial trust structure, resource scale and initial
concentration, and difficulty curves --- whose interactions
characterise the qualitative regimes of the coupled model. The progress-studies reading of
Section~\ref{sec:discussion} closes the paper: the programme is
\emph{prescriptive} as well as descriptive, because the controls it
sweeps are the levers a funding agency actually pulls.
