% =====================================================================
\section{Related Work}
\label{sec:related}
% =====================================================================

The model sits at the intersection of three literatures.

\paragraph{Active inference: cost of mind-change, epistemic
communities, social power.}
\citet{hylandAlbarracin2025MindChange} re-foreground the complexity
term of variational free energy as a motivational quantity: an agent
experiences a cost of revising its own beliefs, proportional to a
Kullback--Leibler divergence between a candidate posterior and the
current one. Because that divergence is asymmetric, belief revision is
path-dependent. We adopt this reading and embed it in the multi-agent
setting: the per-agent cost of mind-change $\stub$ scales a
socially-induced affective utility $\utility$ whose content is the
resource topology. \citet{albarracin2022EpistemicCommunities} establish
the apparatus for coupling active-inference agents through shared
observations and characterise echo-chamber bifurcation as a function
of inter-agent precision; we inherit the coupling apparatus, but treat
the inter-agent precision as itself \emph{learned}, by Gamma-conjugate
precision learning from prediction error.
\citet{albarracin2025SocialPower} formalise social power as the
differential capacity to set policy precision in others' generative
models; the patronage matrix $\flowmat$ in our model is the
operational version of that capacity, made specific to the
resource-allocation setting.

\paragraph{Bayesian rule-induction baseline.}
\citet{oldenburgZhiXuan2024BayesianRule} treat norms as discrete rules
induced from observation via Bayes' rule. Their inference is symmetric
in time: the evidence accrual that locks in a rule will, reversed,
unlock it on a comparable timescale, because there is no learned
social channel, no revision cost, and --- critically --- no resource
topology. This is the natural \emph{trust-and-resource-free} baseline
for our model. The contrast our experimental programme
(Section~\ref{sec:experiments}) targets is between that symmetric
baseline and the asymmetric, path-dependent dynamics that the coupled
belief--resource substrate produces on a structured population.

\paragraph{Sociology of science: field, capital, gatekeeping.}
\citet{bourdieu1975ScientificField} characterises the scientific field
as anisotropic and self-reinforcing: gatekeepers control the
resources that junior agents need to continue investigating, and the
gatekeepers' allocations are shaped by their own paradigm
commitments. This is the empirical phenomenon our resource topology
formalises. The Kuhnian observation \citep{kuhn1962Structure} that
paradigm persistence is sociologically generated, not merely
cognitively maintained, becomes a quantitative claim about the trust
matrix $\trustmat$ and its update rule (Eq.~\ref{eq:trust-update}),
from which the patronage matrix $\flowmat$ is read off
(Eq.~\ref{eq:flow-from-trust}).

\paragraph{Science of science and progress studies.}
The empirical \emph{science of science} literature
\citep{wang2021ScienceOfScience, azoulay2019StarsShadow} documents the
patterns our model is built to generate --- entrenchment thresholds,
gatekeeper effects, the productivity-shadow of star scientists --- but
without a generative substrate against which mechanisms can be tested.
The \emph{progress studies} programme \citep{collison2019ProgressStudies}
asks the prescriptive question --- which funding regimes shorten the
lag between explanatory power and adoption --- but lacks a mechanistic
account of how funding regimes translate into belief dynamics. Our
contribution to both is a substrate-level model in which the funding
ecology's controls (the magnitude of the exogenous inflow
$\inflowrate$ and the concentration of the initial trust matrix
$\trustmat^{(0)}$) feed directly into the belief dynamics that
generate the empirical observables, closing the loop between
prescriptive theory and mechanistic dynamics.

\paragraph{Planck's principle as a limit case.}
\citet{planck1949Autobiography} observed that ``a new scientific truth
does not triumph by convincing its opponents \dots\ but rather because
its opponents eventually die.'' This is the limit case of the
resource-topology argument: when the trust matrix is so entrenched
that no realistic sequence of cross-surprisals can move it, only the
literal turnover of agents can change the gatekeeper distribution. Our model captures the
precursor stage --- resource flow shifting against the old guard ---
without modelling agent birth and death, which we mark explicitly as
out of scope (Section~\ref{sec:discussion}) and as the natural next
extension.
