% =====================================================================
\section{Formal Background}
\label{sec:formal}
% =====================================================================

This section develops the four formal commitments on which the
multi-agent model of Section~\ref{sec:model} rests:
(i)~variational free energy as the agent's objective;
(ii)~the complexity term re-read as a motivational cost of mind-change;
(iii)~precision as attentional weighting on evidence channels; and
(iv)~norms as salience priors --- high-precision components of $\C$
that direct attention to particular features of the environment.

% --------------------------------------------------------------------
\subsection{Variational Free Energy}
\label{ssec:vfe}
% --------------------------------------------------------------------

An active-inference agent carries a generative model
$\Pgen(\obs, \hidden)$ over observations $\obs \in \mathcal{O}$ and
hidden states $\hidden \in \mathcal{S}$, together with a recognition
density $\Q(\hidden)$ that approximates the posterior $\Pgen(\hidden \mid \obs)$.
The agent updates $\Q$ by minimising the variational free energy
\begin{equation}
\FE[\Q] \;=\; \Eunder{\Q}{\ln \Q(\hidden) - \ln \Pgen(\obs, \hidden)}
\;=\; \underbrace{\KL{\Q(\hidden)}{\Pgen(\hidden)}}_{\text{complexity}}
\;-\; \underbrace{\Eunder{\Q}{\ln \Pgen(\obs \mid \hidden)}}_{\text{accuracy}}.
\label{eq:vfe}
\end{equation}
Action selection in active inference proceeds by minimising the
expected free energy of policies; for our purposes,
Eq.~\eqref{eq:vfe} is the only quantity required, because the
model dynamics we develop concern belief revision rather than
multi-step planning.

Prior preferences are encoded as a log-density $\C \in \Real^{|\mathcal{O}|}$
over outcomes, normalised so that $\Pgen(\obs) \propto \exp \C(\obs)$.
The same formal object that other treatments call \emph{values}
\citep{hesp2021DeeplyFelt,torresan2025ShapeC} thus appears here as
$\C$. We will use the two terms interchangeably; the substantive
work is done by what we say about how $\C$ is updated.

% --------------------------------------------------------------------
\subsection{The Complexity Term as a Cost of Mind-Change}
\label{ssec:cost-of-mind-change}
% --------------------------------------------------------------------

Standard treatments of Eq.~\eqref{eq:vfe} read complexity as a
regulariser: a quantity that prevents over-fitting by penalising
posteriors that wander far from the prior. \citet{hylandAlbarracin2025MindChange}
argue that this reading is incomplete. The same KL term, when applied
to the update of an agent's prior preferences from $\C^{(t)}$ to
$\C^{(t+1)}$ in light of new evidence, is most naturally read as
a \emph{motivational cost}: the agent experiences resistance to
revising its prior proportional to
$\KL{\Pgen(\obs \mid \C^{(t+1)})}{\Pgen(\obs \mid \C^{(t)})}$.

Two observations follow. First, this cost is asymmetric in
$\C^{(t)}$ and $\C^{(t+1)}$. The Kullback--Leibler divergence is not a
metric; under a fixed evidence stream, the path from prior $A$ to prior
$B$ is not the time-reverse of the path from $B$ to $A$. Second, the
asymmetry has dynamical consequences whenever the environment is not
stationary. An agent whose update is gated by a positive cost of
mind-change exhibits hysteresis: a cycle of environmental change
followed by reversion to the original state does not return $\C$ to
its starting value. This is the formal signature we exploit in
Section~\ref{sec:model}.

We write the variational $\C$-update as
\begin{equation}
\C^{(t+1)}
\;=\;
\arg\min_{\C}
\Bigl[\,
\KL{\Q^{(t+1)}(\obs)}{\Pgen(\obs \mid \C)}
\;+\;
\stub \,\KL{\Pgen(\obs \mid \C)}{\Pgen(\obs \mid \C^{(t)})}
\,\Bigr],
\label{eq:c-update}
\end{equation}
where $\Q^{(t+1)}(\obs)$ is the posterior over outcomes implied by the
agent's current beliefs about the world, and $\stub \geq 0$ is the
\emph{stubbornness} parameter. Setting $\stub = 0$ recovers the
Bayes-optimal projection of the posterior onto $\Pgen(\obs \mid \C)$;
$\stub \to \infty$ freezes $\C$ at $\C^{(t)}$. Eq.~\eqref{eq:c-update}
is the technical content of the cost-of-mind-change reading;
\citet{hylandAlbarracin2025MindChange} establish it for a single
agent.

\begin{remark}
The asymmetric form
$\KL{\Pgen(\obs \mid \C)}{\Pgen(\obs \mid \C^{(t)})}$
is essential. Replacing it with the symmetric Jensen--Shannon
divergence eliminates the path-dependence of belief revision and so
removes the hysteretic dynamics the model is built to exhibit.
\end{remark}

Figure~\ref{fig:landscape} visualises the effect of $\stub$ on the
location of the variational minimum.

\begin{figure}[t]
\centering
\begin{tikzpicture}
\begin{axis}[
  width=0.78\linewidth, height=5.4cm,
  xlabel={candidate $C$},
  ylabel={objective in Eq.~\eqref{eq:c-update}},
  xmin=-3, xmax=3,
  ymin=0,  ymax=4,
  xtick={-2.0, 0.6}, xticklabels={$C^{(t)}$, $C^{*}$},
  ytick=\empty,
  axis lines=left,
  legend style={font=\scriptsize, draw=none,
                at={(0.98,0.98)}, anchor=north east,
                fill=white, fill opacity=0.85, text opacity=1},
  legend cell align=left,
  every axis plot/.append style={very thick}
]
% Accuracy term: KL[q || p(o|C)] — proxied by (C - C*)^2 / 2 with C* = 0.6
\addplot[domain=-3:3, samples=80, color=black!60, dashed]
  {0.5*(x - 0.6)^2};
\addlegendentry{accuracy term}

% Cost of mind-change: KL[p(C) || p(C_old)] — proxied by (C - C_old)^2 / 2, C_old = -2
\addplot[domain=-3:3, samples=80, color=red!60!black, dotted]
  {0.5*(x - (-2.0))^2};
\addlegendentry{cost of mind-change}

% Combined for lambda = 0
\addplot[domain=-3:3, samples=80, color=blue!30]
  {0.5*(x - 0.6)^2};
\addlegendentry{$\stub = 0$}

% lambda = 0.5
\addplot[domain=-3:3, samples=80, color=blue!60]
  {0.5*(x - 0.6)^2 + 0.5*0.5*(x - (-2.0))^2};
\addlegendentry{$\stub = 0.5$}

% lambda = 2.0
\addplot[domain=-3:3, samples=80, color=blue!90!black]
  {0.5*(x - 0.6)^2 + 2.0*0.5*(x - (-2.0))^2};
\addlegendentry{$\stub = 2.0$}

% Mark minima
\node[circle, fill=blue!30, inner sep=1.4pt] at (axis cs:0.6,0)   {};
\node[circle, fill=blue!60, inner sep=1.4pt] at (axis cs:-0.27,0) {};
\node[circle, fill=blue!90!black, inner sep=1.4pt]
  at (axis cs:-1.13,0) {};

\end{axis}
\end{tikzpicture}
\caption{The cost-of-mind-change reading of the complexity term shifts
the variational minimum. Accuracy alone (dashed black) is minimised at
the Bayes-optimal $C^{*}$; the asymmetric KL penalty (dotted red) is
minimised at $C^{(t)}$. The combined objective in
Eq.~\eqref{eq:c-update} interpolates between the two as $\stub$
increases (blue curves), with the minimum --- marked --- pulled away
from $C^{*}$ toward $C^{(t)}$. As $\stub \to \infty$ the agent freezes
at $C^{(t)}$.}
\label{fig:landscape}
\end{figure}

% --------------------------------------------------------------------
\subsection{Precision as Attention}
\label{ssec:precision-attention}
% --------------------------------------------------------------------

Active-inference agents weight evidence from each of their channels by
a precision parameter $\prc$, formally a parameter of the likelihood
$\Pgen(\obs \mid \hidden; \prc)$ controlling how tightly observations
cluster around their generating hidden state.
\citet{feldmanFriston2010Attention} identify this parameter with
\emph{attention}: high-precision channels dominate posterior updates,
low-precision channels barely register. The mathematical content is
standard. For Gaussian likelihoods, the posterior precision after
combining $K$ channels is $\sum_k \prc_k$; the posterior mean is the
precision-weighted mean of the channel signals.

\citet{parrFriston2017Attention} distinguish \emph{attention} from
\emph{salience}. Attention is a posterior quantity: precision-weighting
applied to currently observed signals. Salience is a prior quantity:
the expected information gain associated with sampling from a channel
or feature, evaluated under the agent's current generative model.
The salience landscape thus determines what the agent is \emph{disposed}
to attend to, before any specific observation is made.

\begin{definition}[Salience prior]
\label{def:salience-prior}
A salience prior on a feature $\feature: \mathcal{O} \to \Real$ is a
component of $\C$ that assigns differential preference weight to
outcomes distinguished by $\feature$, thereby raising the expected
information gain associated with sampling channels that report on
$\feature$. We denote the salience operator by $\salience_{\feature}(\C)$.
\end{definition}

The empirical content of Definition~\ref{def:salience-prior} is that
the shape of $\C$ governs not only what the agent prefers but also
\emph{what the agent attends to}: a sharp $\C$ along feature
$\feature$ produces a high-precision attentional channel for
$\feature$-relevant evidence
\citep{parviziWayne2024PreferencesAttention,limanowskiFriston2018Dark}.
The empirical bridge to attention-as-precision is well-established
\citep{feldmanFriston2010Attention,parrFriston2017Attention}; the
empirical bridge to behaviour is the subject of
\citet{torresan2025ShapeC}.

% --------------------------------------------------------------------
\subsection{Norms as Salience Priors}
\label{ssec:norms-as-salience}
% --------------------------------------------------------------------

We are now in a position to state the central conceptual move. A norm,
on the present account, is not a content prior over actions but an
attentional prior over what features of a situation count as
normatively relevant. The honesty norm does not say ``say true things'';
it says ``attend to truth-conditions of statements you make.''
The generosity norm does not say ``give things away''; it says
``attend to others' material needs.'' Norm-following behaviour is
then the policy that minimises expected free energy under the
agent's normative salience landscape, in concert with whatever
non-normative preferences the agent holds.

\begin{definition}[Norm]
\label{def:norm}
A norm $\norm$ is a salience prior $\salience_{\feature_\norm}(\C)$
that is shared across a community $V_\norm \subseteq V$ in the
precise sense that the agents in $V_\norm$ assign correlated
high-precision weight to feature $\feature_\norm$ in their
respective $\C$.
\end{definition}

Two consequences of Definition~\ref{def:norm} should be flagged. First,
norms are objects of \emph{population-level} attention rather than
individual conviction; the binding force of a norm derives from its
being shared, in the sense that an agent's neighbours also weight the
same feature highly. Second, the precision-weighting structure of
norms means that whoever modulates the precision of the channel that
installs a feature into $\C$ thereby modulates which norms become
active in the community. This is the formal bridge to the
multi-agent model of Section~\ref{sec:model}: \emph{trust} ---
the precision an agent assigns to evidence supplied by neighbours ---
is the variable through which norms are installed in
$\C$ and through which they are maintained against environmental drift.
Figure~\ref{fig:variables-loop} summarises the four variables and
their update relations.

\begin{figure}[t]
\centering
\begin{tikzpicture}[
  box/.style    = {draw, rounded corners=3pt, minimum width=3.4cm,
                   minimum height=1.1cm, font=\small, align=center,
                   fill=gray!5},
  varbox/.style = {draw, rounded corners=3pt, minimum width=3.4cm,
                   minimum height=1.1cm, font=\small, align=center,
                   fill=blue!10, draw=blue!60, very thick},
  arr/.style    = {-Stealth, semithick},
  lbl/.style    = {font=\scriptsize, midway, fill=white,
                   inner sep=2pt, align=center}
]
% Four corners of a wide rectangular loop
\node[varbox]                          (Ci)    {\textit{salience prior} $\C_i$};
\node[box, right=4.6cm of Ci]          (act)   {action $a_i$};
\node[box, below=2.2cm of act]         (post)  {posterior $\Q_i$ \\
                                               \scriptsize Eq.~\eqref{eq:posterior-mix}};
\node[varbox, below=2.2cm of Ci]       (trust) {\textit{trust} $\trust{i}{j}$};

% Loop edges, clockwise
\draw[arr] (Ci)    -- node[lbl]{sample from $\C_i$} (act);
\draw[arr] (act)   -- node[lbl]{neighbours observe; \\
                                environment reveals $\hidden^{*}$} (post);
\draw[arr] (post)  -- node[lbl]{prediction error \\
                                Eq.~\eqref{eq:trust-update}} (trust);
\draw[arr] (trust) -- node[lbl]{weights neighbours \\
                                in Eq.~\eqref{eq:posterior-mix}} (Ci);

% Diagonal arrow: posterior drives C-update
\draw[arr, dashed] (post) -- node[lbl, pos=0.42]
  {Eq.~\eqref{eq:c-update-multinorm},~$\stub_{i,r}$} (Ci);
\end{tikzpicture}
\caption{The four variables of the multi-agent model and their update
loop, traversed clockwise. The salience prior $\C_i$ shapes the
agent's action; the action is observed by the social neighbourhood
and the environment reveals ground truth; the resulting posterior
$\Q_i$ both drives an asymmetric-KL update of $\C_i$ (dashed,
gated by stubbornness $\stub_{i,r}$) and supplies the prediction
errors that update channel precision $\trust{i}{j}$. The two
boxed quantities are the dynamical state of agent $i$ in this paper;
the grey nodes are derived per-step quantities.}
\label{fig:variables-loop}
\end{figure}
