% =====================================================================
\section{Experimental Programme}
\label{sec:experiments}
% =====================================================================

The coupled model of Section~\ref{sec:model} has three independently-%
controlled levers: the structure of the initial trust matrix
$\trustmat^{(0)}$ (which determines the initial resource topology
through Eqs.~\ref{eq:flow-from-trust}--\ref{eq:inflow-share}); the
scale of the exogenous resource economy
($\inflowrate$ and decay $\rdecay$) together with how concentrated the
initial trust is across the population; and the coupling between
experiment cost and Fisher information (the resource-gated cost
$\gatecost(\xexp;\resource{i})$ of
Eq.~\ref{eq:expected-resource-gain}). The experimental programme is
organised around three corresponding axes that probe each lever in
isolation and, at their interaction, characterise the qualitative
regimes of the model. Each is reported through per-community and
per-agent observables (Remark~\ref{rem:no-pool}) and through
\emph{resource-weighted} population summaries, never a single pooled
mean.

\paragraph{Axis~A --- Initial trust structure.}
Fix the social graph $\graph$, the total inflow $\inflowrate$, and the
initial belief distribution. Vary the structure of the initial trust
matrix $\trustmat^{(0)}$ from \emph{uniform} (every edge weighted
equally) through \emph{block-clustered} (within-community trust high,
between-community trust low --- the ``schools'' regime) to
\emph{hub-aligned} (a small set of agents receive high incoming trust
from everyone). Run the population through a paradigm-shift schedule
$A \to B$. The hypothesis is that time-to-shift is \emph{non-monotone}
in block-clustering: uniform trust spreads evidence too thinly across
schools and shifts are slow; strong block-clustering makes each school
shift as a unit but blocks transmission \emph{between} schools and
also slows global shifts; an intermediate clustering --- enough
within-school trust to integrate evidence locally, enough
cross-school trust to transmit a vanguard's conversion to the
mainstream --- maximises shift speed. The primary observables are the
time at which resource-weighted bimodality collapses; the identity of
the top-$k$ resource-centrality agents at the start and end of the
run, with their belief alignment to old and new paradigms; and the
within-school vs cross-school components of the trust spectrum
(Fiedler value, see Section~\ref{sec:experiments}) tracked through the
shift event.

\paragraph{Axis~B --- Resource scale and initial concentration.}
Hold $\graph$ and the initial belief distribution fixed. Sweep two
controls jointly: (a)~the total exogenous inflow rate $\inflowrate$
relative to the decay $\rdecay$, which sets the steady-state
population resource scale, and (b)~the concentration of the initial
trust matrix as measured by the Gini of aggregate incoming trust
$\{\sum_j \trust{j}{i}^{(0)}\}_i$, swept from $0$ (uniform: every
agent receives the same exogenous share) to near $1$ (a single hub
absorbs all exogenous inflow). The substantive hypothesis has two
parts. First, the entrenchment threshold above which paradigm shifts
fail rises with initial trust concentration: when one community
already absorbs most of the exogenous inflow, the trust update of
Eq.~\eqref{eq:trust-update} has to overcome a head start in
$\trust{j}{\mathrm{hub}}$ before reallocation can begin. Second,
following the historical observation that scientific revolutions
coincide with expansions of the scientific enterprise rather than
with stable or contracting periods, higher $\inflowrate$ facilitates
shifts because new resources entering can be captured by new
alignment patterns without having to be wrested from old ones.
Observables include those of Axis~A plus the endogenous resource
Gini at steady state and its trajectory across the shift event, and
the correlation between $\resource{i}$ and within-community average
$\surprisal_{ii}$ --- high when the trust update has had time to
re-route resources toward the agents whose predictives now match
reality, low when concentration is locked in.

\paragraph{Axis~C --- Difficulty curves and the fitness-valley effect.}
Use the Fisher-information--coupled experiment cost
$\gatecost(\xexp;\resource{i}) = c_{0}\,\hone(\xexp)^{2}/\sigobs^{2} \cdot
\phi(\resource{i})$ of Section~\ref{ssec:policy}. Sweep the steepness
parameter and the threshold $r_{\min}$ in $\phi$; independently vary the
placement of $\thetastar$ across cheap and expensive $\thetapar$-regions
(where ``cheap'' means $\thetastar$ is well-discriminated by
low-$\hone(\xexp)$ experiments and ``expensive'' means it requires
high-$\hone(\xexp)$ experiments). The hypothesis is that when
$\thetastar$ lies in an expensive region, paradigm shifts are gated by
\emph{resource accumulation} rather than by belief dynamics, with a
sharp discoverability transition at the resource threshold where the
discriminating experiments become accessible. Below threshold the
discriminating evidence is essentially never collected and any
apparent shift is the result of agents arriving at the right answer
for the wrong reasons; above threshold the discriminating evidence is
collected and the standard belief-dynamical mechanisms take over.
This connects to the historical observation that paradigm shifts
requiring genuinely new experimental capabilities (relativity, quantum
mechanics) take far longer to resolve than those settleable with
existing apparatus.

\paragraph{The interaction grid.}
The three axes give the basic decomposition. Their interactions are
where the qualitative regimes live, and the programme explicitly
probes them. Axis~A$\times$B: block-clustered trust matters more
when the population resource scale is small (every school must fund
its own apparatus), and matters less in the high-$\inflowrate$ regime
where each school's vanguard can be supported by the trust-aggregated
exogenous inflow. Axis~A$\times$C: a vanguard with the resources to
run expensive experiments converts the mainstream only if the
cross-school trust pathway from vanguard to mainstream is
sufficiently broad. The deepest interaction is B$\times$C:
high-concentration regimes can in principle fund expensive
experiments, but only if the high-trust hubs already credit the
discriminating-regime paradigm, which by Eq.~\eqref{eq:trust-update}
they do not when their cross-surprisals on the new paradigm are still
large; concentrated regimes can therefore systematically suppress
discriminating work, while uniform regimes diffuse resources too
thinly for individual agents to afford expensive apparatus, with an
intermediate concentration --- both concentrated enough to fund
apparatus and distributed enough that the apparatus is built when
needed --- as the conjectured ``productive'' regime of normal science.
We identify three representative settings on each axis and run the
$3\times3\times3$ interaction grid with multiple seeds per cell.

\paragraph{What gets measured.}
Across all experiments the primary observables are:
(i)~the resource-weighted population belief distribution over time,
including its mean, variance, and bimodality;
(ii)~time-to-shift, defined as the time at which the
resource-weighted bimodality collapses or the resource-weighted mean
crosses a paradigm threshold;
(iii)~the identity of the top-$k$ resource-centrality agents at
start and end of the run, with their belief alignment to old and new
paradigms;
(iv)~the correlation between $\resource{i}$ and belief alignment with
current gatekeepers, tracked through shift events;
(v)~reachability and discoverability for Axis~C; and
(vi)~the endogenous resource Gini at steady state.
These observables are designed so that each can be plotted against
each axis sweep, and so that the interaction effects can be
characterised in two-dimensional plots without further derivation.

\begin{figure}[!ht]
\centering
\begin{tikzpicture}
\begin{axis}[
  width=0.78\linewidth, height=5.0cm,
  xlabel={environmental truth $\thetastar(t)$},
  ylabel={per-community mean $\bar\muest_{\community{k}}$},
  xmin=-0.15, xmax=1.15,
  ymin=-0.15, ymax=1.15,
  xtick={0,1}, xticklabels={$A$, $B$},
  ytick={0,1}, yticklabels={$\thetapar_A$, $\thetapar_B$},
  axis lines=left,
  legend style={font=\scriptsize, draw=none,
                at={(0.02,0.98)}, anchor=north west,
                fill=white, fill opacity=0.85, text opacity=1},
  legend cell align=left,
  every axis plot/.append style={very thick}
]
\addplot[domain=0:1, samples=40, color=gray!55, dashed] {x};
\addlegendentry{decoupled baseline (no loop)}
\addplot[domain=0:1, samples=80, color=blue!60, smooth]
  ({x}, {x - 0.12*sin(deg(pi*x))});
\addlegendentry{vanguard $\community{1}$ (high $\resource{}$)}
\addplot[domain=0:1, samples=80, color=blue!60, smooth, forget plot]
  ({1-x}, {(1-x) + 0.12*sin(deg(pi*(1-x)))});
\addplot[domain=0:1, samples=80, color=red!70!black, smooth]
  ({x}, {x - 0.34*sin(deg(pi*x))});
\addlegendentry{mainstream $\community{2}$ (low $\resource{}$)}
\addplot[domain=0:1, samples=80, color=red!70!black, smooth, forget plot]
  ({1-x}, {(1-x) + 0.34*sin(deg(pi*(1-x)))});
\draw[->, thick, red!70!black] (axis cs:0.55,0.37) -- (axis cs:0.65,0.44);
\draw[->, thick, red!70!black] (axis cs:0.45,0.63) -- (axis cs:0.35,0.56);
\node[font=\scriptsize, red!70!black] at (axis cs:0.80,0.28) {forward};
\node[font=\scriptsize, red!70!black] at (axis cs:0.20,0.72) {reverse};
\node[circle, fill=black, inner sep=1.2pt] at (axis cs:0,0) {};
\node[circle, fill=black, inner sep=1.2pt] at (axis cs:1,1) {};
\end{axis}
\end{tikzpicture}
\caption{Predicted per-community hysteresis under the reversal schedule
of Claim~\ref{claim:path-dependence}, at a representative cell of the
$3\times3\times3$ interaction grid (block-clustered initial trust,
high initial-trust concentration, expensive-region $\thetastar$). A
vanguard community ($\community{1}$, blue) with sufficient resources
to run discriminating experiments converts quickly and encloses a
small loop; a mainstream community ($\community{2}$, red) held back
by a resource bottleneck encloses a large one. The decoupled baseline
($\gatecost$ independent of $\resource{i}$ and $\betaexp$ taken to
the high-temperature limit so that
$\E[\Delta\resource{i}\!\mid\!\xexp]$ does not differentiate
experiments) collapses to the identity line in the homogeneous limit.
Plotting per community --- rather than a single resource-weighted
mean --- is what keeps the contrast visible (Remark~\ref{rem:no-pool}).}
\label{fig:reversal-loop}
\end{figure}

\noindent The full $3\times3\times3$ grid, multi-seed runs, and
quantitative phase diagrams are deferred to a companion treatment;
this paper establishes the programme, the predicted regimes, and a
gating result from a minimal two-community simulation that confirms
the resource-coupled substrate produces qualitatively different
shift-dynamics under uniform versus concentrated exogenous inflow.
