% =====================================================================
\section{Discussion}
\label{sec:discussion}
% =====================================================================

The model articulates paradigm persistence as the dynamics of a
\emph{coupled belief--resource system} organised around one Bayesian
learned relational object: the trust matrix $\trustmat$. The patronage
flow $\flowmat$ and the per-agent resource state $\resource{i}$ are
deterministic readouts of $\trustmat$
(Eqs.~\ref{eq:flow-from-trust}--\ref{eq:inflow-share}); the
cost-of-mind-change $\stub$ is a per-agent cognitive trait that scales
an affective preference for posteriors aligned with resource-rich
neighbours (Eq.~\ref{eq:utility}). What would otherwise read as two
qualitatively distinct funding channels --- a ``patronage'' channel
allocated by social standing and a ``meritocratic'' channel allocated
by predictive accuracy --- collapses here into one quantity. The
substantive claim is that the collapse is correct: in real scientific
institutions trust \emph{is} what social standing is, when standing is
read off the community's posterior on each agent's predictive
reliability. The classical concern --- that a pure-patronage model
predicts any entrenched paradigm stays entrenched forever --- does not
apply, because $\trustmat$ updates by Bayesian precision learning
(Eq.~\ref{eq:trust-update}) on cross-surprisals: superior predictive
accuracy translates into superior resource access through the same
trust matrix that allocates resources. The price of the collapse is
also its content: gatekeepers and the field's reviewers are not
modelled as institutionally separate actors with separable utilities,
but as readouts of one underlying trust topology at two aggregation
scales.

\paragraph{A progress-studies reading.}
Axis~B of Section~\ref{sec:experiments} makes the model
\emph{prescriptive}: the magnitude of the exogenous inflow
$\inflowrate$ and the concentration of the initial trust matrix
$\trustmat^{(0)}$ are the levers a funding ecology controls. The
model generates concrete hypotheses that empirical science-of-science
can test \citep{wang2021ScienceOfScience, azoulay2019StarsShadow,
collison2019ProgressStudies}: (i)~holding total funding fixed,
decreasing the concentration of incoming trust lowers entrenchment
thresholds, an effect that should be visible in natural experiments
on grant-policy reforms that broaden the set of effective gatekeepers
without changing the budget; (ii)~the productive funding regime is
intermediate in trust concentration, not at either extreme, testable
against cross-field comparisons of paradigm-shift latency;
(iii)~scientific revolutions coincide with $\inflowrate$ expansion
rather than with stable or contracting periods, because new resources
can be captured by new trust patterns without having to be wrested
from old ones. The mechanistic active-inference account supplies a
substrate against which these hypotheses become quantitative
predictions rather than post-hoc rationalisations of historical case
studies.

\paragraph{Out of scope.}
Three cuts are explicit. First, \emph{agent birth and death}: Planck's
remark that ``a new scientific truth does not triumph by convincing
its opponents \dots\ but rather because its opponents eventually die''
\citep{planck1949Autobiography} is the limit case of the resource-%
topology argument, captured here only at the precursor stage
(resource flow shifting against the old guard as $\trustmat$ updates
under sustained cross-surprisals). The relative timescales of belief
revision and generational replacement, and the dynamics of
graduate-student spawning by gatekeepers, are the natural next
extension. Second, \emph{endogenous graph formation}: the social
graph $\graph$ is exogenous by design --- this paper aims to isolate
the dynamics of trust and belief on a fixed structure, not to explain
the structure itself. Allowing $\graph$ to form from $\trustmat$ is an
obvious follow-up but introduces a dynamical claim we are not making
here. Third, an \emph{information-geometric reading of power}: the
trust-derived flow $\flowmat$ is, on the gauge-field reading of
social power developed in \citep{albarracin2025SocialPower}, the
operational signature of who sources the cost-field for whom. A
formal treatment requires anisotropic-metric machinery beyond the
present scope, and is the natural theoretical home of the coupled
model.

The unification of patronage and meritocratic allocation through
$\trustmat$ is itself a modelling commitment, not a hidden parameter
of the experimental programme. The substantive future-work question it
opens is the \emph{temporal evolution of the trust-allocation
mapping}: Eqs.~\eqref{eq:flow-from-trust}--\eqref{eq:inflow-share}
treat the readout from $\trustmat$ to $\flowmat$ and $\inflowshare{i}$
as fixed, but in reality the institutional rules that translate peer
credit into resource allocation respond to perceived crises and
reform episodes. Modelling this meta-level dynamics turns the model
from a theory of paradigm persistence into a theory of the conditions
under which scientific institutions reorganise the trust-to-resource
mapping itself to escape paradigm capture.
