% Section: §5 Structural Extension --- Bayes Nets and Bayesian Model Reduction
% PDF context: main technical section. Each agent's paradigm becomes a Bayesian
% network with a core/belt partition (Lakatos), structural revisions are scored
% by Bayesian model reduction (Friston-Penny 2011; Friston-Parr-Zeidman 2018),
% and observation is recast as a precision-budget allocation over belt nodes
% governed by expected free energy (Mirza, Parr-Friston).
% Drawing from: cluster_C3_bmr_lakatos (BMR + Lakatos + Thagard/Quine ancestry),
% cluster_C7_attention_precision (active sampling under EFE), and master refs.bib
% for the active-inference scaffolding.
% Entries cited: 11
% Full-text papers consulted:
%   Friston2018_2  -- Friston, Parr, Zeidman, "Bayesian model reduction" (technical anchor for reduced free energy + Dirichlet/Gaussian cases)
%   Mirza2019_2    -- Mirza, Adams, Friston, Parr, "Bayesian model of selective attention" (closing precision-allocation move)
% Key renaming TODOs (Semantic Scholar auto-keys -> hand-curated):
%   Friston2018_2     -> fristonParrZeidman2018BMR             % Anchor: BMR technical review (reduced free energy across exponential family)
%   Thagard1989_1     -> thagard1989ExplanatoryCoherence       % Anchor: ECHO / explanatory coherence canonical statement (BBS target article)
%   Thagard1988_4     -> thagardNowak1988ContinentalDrift      % Anchor: ECHO applied to a Kuhnian paradigm shift
%   Ranney1988_3      -> ranneyThagard1988NaivePhysics         % Connectionist belief-revision precedent
%   Markovi'c2023_1   -> markovicFristonKiebel2024BMRSparsify  % BMR applied to deep-network sparsification (modern follow-up)
%   Heijdra1986_6     -> heijdraLowenberg1986DuhemQuineLakatos % Duhem-Quine + Lakatos joint reading
%   Wahyudi2026_2     -> wahyudi2026QuineWebOfBelief           % Quine web-of-belief as coherentist justification
%   Mirza2019_2       -> mirzaAdamsFristonParr2019SelectiveAtt % Anchor: precision-as-attention in MDP active inference
% TODO refs.bib additions (foundational anchors that neither cluster surfaced):
%   lakatos1970FalsificationMethodology -- Lakatos, "Falsification and the methodology of scientific research programmes", in Lakatos & Musgrave (eds.), 1970 (the core/belt locus classicus)
%   quine1951TwoDogmas                  -- Quine, "Two dogmas of empiricism", Philosophical Review 60(1), 1951 (web-of-belief / underdetermination)
%   fristonPenny2011PostHocBMS          -- Friston & Penny, "Post hoc Bayesian model selection", NeuroImage 56(4), 2011 (variational-Laplace BMR derivation; PDF cites this as the technical foundation)
%   thagardVerbeurgt1998CoherenceConstraint -- Thagard & Verbeurgt, "Coherence as constraint satisfaction", Cognitive Science 22(1), 1998 (formal ECHO algorithm)

\section{Structural Extension: Bayes Nets and Bayesian Model Reduction}
\label{sec:structural}

The scalar treatment of \S\ref{sec:scalar} sufficed to expose
asymmetric belief-revision dynamics, but it cannot speak to what
practising scientists actually argue about: which commitments are
foundational, which are negotiable, and which channel of observation
to spend a finite attentional budget on next. The structured extension
recovers all three by promoting each agent's paradigm from a point in
$\Theta$ to a Bayesian network whose nodes carry theoretical
commitments and whose edges encode inferential dependencies.

\paragraph{Structural belief representation.}
The conceptual ancestry of paradigms-as-networks predates active
inference by half a century. \citet{Wahyudi2026_2}, reconstructing
Quine's coherentism, summarises the web-of-belief picture in which no
proposition confronts experience in isolation: revisions propagate
along inferential links, and the choice of which link to cut is
underdetermined by data alone. Lakatos's methodology of scientific
research programmes sharpens this into a topology with a designated
\emph{hard core} of commitments shielded by a flexible
\emph{protective belt} of auxiliary hypotheses
\citep{Heijdra1986_6,Ostapiuk2025_11}; the belt absorbs anomalies so
that the core remains in force, and a programme degenerates only when
belt-patching ceases to predict novel facts. Thagard's ECHO programme
operationalises Quinean coherence as a connectionist constraint
network and demonstrates, on the continental-drift case study, that
the same machinery can reconstruct a historical paradigm shift as a
re-equilibration of explanatory links
\citep{Thagard1989_1,Thagard1988_4}; \citet{Ranney1988_3} extend the
account to belief revision in naive physics. None of these accounts
furnishes a generative model, but each picks out the same structural
object: a graph of commitments whose densely-connected interior is
held rigid while its periphery negotiates with observation. We adopt
this object verbatim, identifying core nodes with the parameters
under tight prior precision and belt nodes with those under broad
prior precision in the agent's generative model.

\paragraph{Bayesian model reduction.}
Promoting the core/belt distinction to a piece of formal machinery
requires a principled treatment of \emph{structural} inference --- the
problem of scoring alternative graphs, not just alternative parameter
values, against the same data. \citet{Friston2018_2}, extending the
post-hoc model selection scheme of Friston \& Penny (2011), provide
precisely this in the form of Bayesian model reduction (BMR). The
construction is straightforward in variational Bayes: once the full
model has been inverted, the evidence and posterior of any
\emph{reduced} model --- one differing only in its priors --- can be
computed analytically from the full posterior and the ratio of priors,
generalising the Savage--Dickey density ratio. Concretely, switching
off a candidate edge corresponds to collapsing the prior variance on
the associated parameter to zero; the resulting change in variational
free energy is the (log) Bayes factor for that structural revision.
For the Dirichlet posteriors that govern categorical likelihood
mappings in discrete state-space models --- the natural form for
theory--observation links --- the reduced free energy has a closed
form in terms of multivariate beta functions. \citet{Friston2018_2}
showcase BMR as both a data-analytic tool and a metaphor for
neurobiological abductive reasoning: the same equations that prune
redundant DCM connections describe the synaptic elimination an agent
performs when ``thinking off-line'' simplifies its model of a
contingency. \citet{Markovi'c2023_1} extend the scheme to deep-network
sparsification, evidencing that the reduction is general across
parameter spaces. We invoke BMR as the engine of paradigm dynamics:
candidate revisions to an agent's network --- adding a belt link,
demoting a core node, accepting a rival auxiliary --- are scored by
their reduced free energy, with the structural affective utility
inherited from \S\ref{sec:motivatedbelief} entering as a contribution
to the cost of admitting each revision. Path-dependence in the scalar
model thus refines into something graph-theoretic: revisions that
touch the core are more expensive because more of the posterior has
to be re-balanced.

\paragraph{Active sampling and precision allocation.}
The structural representation closes the loop with action selection in
a way the scalar model could not. Observation in the structured model
is not exogenous arrival of evidence but an active sampling decision:
the agent allocates a finite precision budget across belt nodes,
sampling some channels with high precision and others with low.
\citet{friston2015ActiveInferenceEpistemic} establishes the
underlying primitive --- expected free energy decomposes into
extrinsic (utility) and epistemic (information-gain) value ---
and \citet{millidge2021EFE} surveys the equivalent decompositions.
\citet{Mirza2019_2} show, in an MDP formulation of Yarbus's classic
visual-search paradigm, that selective attention is just
context-dependent modulation of the likelihood precisions $\zeta$:
raising $\zeta$ on task-relevant channels and lowering it on
task-irrelevant ones reweights expected information gain so that the
agent saccades to evidence that distinguishes hypotheses it actually
entertains. The same construction underwrites
\citet{feldmanFriston2010Attention}, \citet{parrFriston2017Attention},
and the precision-control account of
\citet{parrFriston2019AdvancedAtt}; \citet{limanowskiFriston2018Dark}
fills in the failure mode where mis-allocated precision dampens
exploration. Transposed to the paradigm setting, precision allocation
\emph{is} the paradigm-dependence of observation: an agent inhabiting
a Newtonian network attends to mass-and-force belt nodes with high
precision and assigns negligible precision to channels whose only
hidden cause lies in a rival network. The result is a structured
analogue of theory-laden perception
(\S\ref{sec:theoryladen})~\citep{hylandAlbarracin2025MindChange},
now derived rather than stipulated: what counts as a salient
observation is what the agent's current network has bandwidth to
believe.
